----
title: "Data Analysis"
Author: "Maheep Mundra, Addison Dinglasan"
Date: "2026-02-23"
output: html_document
vignette: >
  %\VignetteIndexEntry{Lizard Data Analysis Vignette}
  %\VignetteEngine{knitr:rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: bibliography.bib
nocite: "@*"
----
# Import libraries and methods
```{r}
library(tidyverse)
library(ggplot2)
library(knitr)
source("submissions/BasicImplementation/bootstrap.R")
```
# Load Data
```{r}
lizard_data <- read.csv("submissions/BasicImplementation/data.csv",
 fileEncoding = "latin1")

# Inspect data
print(head(lizard_data, 5))
```
# Clean up data
```{r}
#Clean names replace . with _ between word spaces,
#remove white spaces and lowercase
colnames(lizard_data) <- tolower(gsub(" ", "_", trimws(gsub("\\.+", " ", colnames(lizard_data)))))

#Print the first 50 entries of dataset
print(head(lizard_data, 50))
#Print Size of data set
print(dim(lizard_data))

#Remove rows with missing sex, snout_vent_length_mm or body_temperature_c
clean_data <- lizard_data %>% 
    filter(
    !is.na(sex),
    !is.na(snout_vent_length_mm),
    !is.na(body_temperature_c))

#Print the first 5 entries of dataset
print(head(clean_data, 5))
#Print Size of data set
print(dim(clean_data))
```
# Extract data
```{r}
# Snout Length values
male_snout_len <- clean_data[    # Male snout length
    clean_data["sex"] == "M",
    "snout_vent_length_mm"]

female_snout_len <- clean_data[  # Female snout length
    clean_data["sex"] == "F",
    "snout_vent_length_mm"]

# Body Temps
f_gravid_btemp <- clean_data[     # Body temp of pregnant females
    clean_data["reproductive_status"] == "gravid",
    "body_temperature_c"]

f_nongravid_btemp <- clean_data[  # Body temp of non-pregnant females
    clean_data["reproductive_status"] == "nongravid",
    "body_temperature_c"]
```

# Apply two sample bootstrap function to calculate mean
```{r}

#Bootstrap estimate for mean Snout Lengths
boot_mean_snout_len <- two_sample_bootstrap(
    male_snout_len, 
    female_snout_len,
    iterations = 1000,
    stat = "mean")

print(head(boot_mean_snout_len, 5))

# Bootstrap estimate for mean Body temp of gravid and non-gravid females
boot_mean_body_temp <- two_sample_bootstrap(
    f_gravid_btemp, 
    f_nongravid_btemp,
    iterations = 1000,
    stat = "mean")


```
# Generate Bootstrap Distribution Histogram and Comparative Statistics Table for Bootstrap estimate for mean Snout Lengths (Male vs Female)
```{r}
# Convert to dataframe
boot_mean_snout_len_dist <- data.frame(diff_len = boot_mean_snout_len)

#Plot bootstrap Distribution Histogram
ggplot(boot_mean_snout_len_dist, aes(x = diff_len))  +
    geom_histogram(bins = 30, fill = "lightblue", color = "black") +
    geom_vline(xintercept = mean(boot_mean_snout_len),
             color = "red", size = 1.2) +
  labs(
    title = "Bootstrap Distribution: Snouth length Difference (Male vs Female)",
    x = "Difference in Mean snout lengths (mm)",
    y = "Frequency"
  )
 #Comparative Table
comparative_table_for_mean_len <- data.frame(
  Statistic = c("Mean", "Median", "Standard Deviation"),
  Male = c(
    mean(male_snout_len),
    median(male_snout_len),
    sd(male_snout_len)),
  Female = c(
    mean(female_snout_len),
    median(female_snout_len),
    sd(female_snout_len)
  ))

# Print table
print(comparative_table_for_mean_len)
```
  # Generate Bootstrap Distribution Histogram and Comparative Statistics Table for Bootstrap estimate for mean Body temperatures (Gravid vs Non-Gravid)
```{r}
# Convert to dataframe
boot_mean_body_temp_dist <- data.frame(diff_temp = boot_mean_body_temp)

#Plot bootstrap Distribution Histogram
ggplot(boot_mean_body_temp_dist, aes(x = diff_temp))  +
    geom_histogram(bins = 30, fill = "darkblue", color = "black")+
    geom_vline(xintercept = mean(boot_mean_body_temp),
             color = "red", size = 1.2) +
  labs(
    title = "Bootstrap Distribution: Body Temprature Difference (Gravid vs Non-Gravid)",
    x = "Difference in Body Temprature (celsius)",
    title = "Bootstrap Distribution: Body Temperature Difference (Gravid vs Non-Gravid)",
    x = "Difference in Mean snout lengths (mm)",
    y = "Frequency"
  )
 #Comparative Table
comparative_table_for_mean_body_temp <- data.frame(
  Statistic = c("Mean", "Median", "Standard Deviation"),
  Gravid = c(
    mean(f_gravid_btemp),
    median(f_gravid_btemp),
    sd(f_gravid_btemp)),
  Nongravid = c(
    mean(f_nongravid_btemp),
    median(f_nongravid_btemp),
    sd(f_nongravid_btemp)
  ))

# Print table
print(comparative_table_for_mean_body_temp)
```
# AI Declaration 

The authors used Gemini and ChatGPT for technical troubleshooting/debugging R error messages and  Git version control conflicts. The models also helped generate ideas for test cases and provided R grammar suggestions and feedback on formatting. The authors reviewed any AI-suggested logic and feedback. 

# Bibliography

